---
title: Start temperature does not affect copepod CTmax
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
bibliography: references.bib
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

Several studies have shown that the temperature a CTmax assay starts at can affect the observed upper thermal limit [@terblanche2007]. This seems like it may be species-specific, however, with highly variable results across species [@faber], and even across populations of the same species [@cicchino2024].

While the effect of start temperature can be quite clear, the underlying mechanism is still unknown. Much of the past work has been done with terrestrial organisms (like insects) - in these systems, dehydration is a major risk during experiments. As a result, the effect of starting temperature on CTmax may reflect the changes in duration, rather than the direct effect of a different initial temperature.

In theory, without confounding effects of other processes like dehydration and starvation, starting temperature should have minimal effect on CTmax so long as the assay starts below the critical temperature, or the point where damage accumulation outpaces damage repair [@faber]. This project examines how CTmax varies with starting temperature in Skistodiaptomus pallidus, a widespread freshwater calanoid copepod.

## Methods

We ran CTmax assays starting at different temperatures for several species, spanning a range of prior conditions.

| Species                   | Acclimation History                                     | Acclimation Temperature     |
|------------------|-----------------------------------|-------------------|
| Skistodiaptomus pallidus  | 10 generations in laboratory culture                    | 16°C                        |
| Leptodiaptomus siciloides | 1 week under laboratory conditions                      | 23\*C                       |
| Onychodiaptomus birgei    | Field collected (\<5 hours under laboratory conditions) | Variable field temperatures |

We followed a standard protocol for measuring CTmax in copepods. Briefly, a small water bath was filled with 2 L of water and adjusted to the desired temperature either by adding ice or with a small aquarium heater. Once at the correct temperature, copepods were added individually into 50 mL glass vials with 10 mL of artificial freshwater medium. Vials were held in the water bath and the water was already equilibrated to the bath temperature. As such, the transition from holding temperature to starting temperature was acute.

## Results

### Water temperatures and ramping rates

We tracked temperature in the water bath during one replicate for each starting temperature to ensure that ramping rates behaved as expected. Temperature loggers were placed in each water bath and recorded temperature every 30 seconds (every 1 minute in some cases).

```{r}
ggplot(comb_data, aes(x = time_point, y = temp_c, colour = start_temp)) + 
  geom_point() + 
  labs(y = "Water Temp. (°C)",
       colour = "Start Temp. (°C)", 
       x = "Assay Time (minutes)") + 
    theme_matt() + 
  theme(legend.position = "right")
```

Ramping rates (calculated for 5 minute intervals throughout the experiment) decreased over time, as expected.

```{r}

mean_temps = comb_data %>% 
  mutate(minutes = time_point) %>% 
  drop_na(temp_c) %>% 
  group_by(start_temp, ten_min_int) %>% 
  summarise(mean_temp = mean(temp_c))

ramp_rates = comb_data %>% 
  mutate(minutes = time_point) %>% 
  drop_na(temp_c) %>% 
  group_by(start_temp, ten_min_int) %>% 
  nest() %>%
  mutate(
    # Fit the linear model to each interval's data
    model = map(data, ~ lm(temp_c ~ minutes, data = .x)),
    # Tidy the model object into a dataframe
    tidied = map(model, tidy)
  ) %>%
  unnest(tidied) %>%
  # Filter to get just the slope (the coefficient for 'time')
  filter(term == "minutes") %>% 
  inner_join(mean_temps, by = c("start_temp", "ten_min_int"))


ggplot(ramp_rates, aes(x = ten_min_int, y = estimate, colour = start_temp)) + 
  geom_line(linewidth = 2) + 
  labs(y = "Ramp Rate (°C per minute)",
       x = "Ten Minute Interval", 
       colour = "Start Temp. (°C)") + 
    theme_matt() + 
  theme(legend.position = "right")
```

As shown in previous work with this setup, this decrease in ramping rates was strongly related to the average water temperature during the time interval, regardless of starting temperature.

```{r}

ggplot(ramp_rates, aes(x = mean_temp, y = estimate, colour = start_temp)) + 
  geom_line(linewidth = 2) + 
  labs(y = "Ramp Rate (°C per minute)",
       x = "Average Temp. (°C)", 
       colour = "Start Temp. (°C)") + 
  theme_matt() + 
  theme(legend.position = "right")

```

## CTmax Data

There was no relationship between observed CTmax and the starting temperature.

```{r}

ggplot(trait_data, aes(x = start_temp, y = ctmax)) + 
  facet_grid(species~.) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm") + 
  labs(y = "CTmax (°C)",
       x = "Start Temp. (°C)") + 
  theme_matt_facets()

```

This is supported by the results of a linear regression, which shows no relationship between CTmax and starting temperature.

```{r}

model_data = trait_data %>% 
  mutate(ctmax_cent = scale(ctmax, center = T, scale = F),
         start_cent = scale(start_temp, center = T, scale = F))

temp.model = lme4::lmer(data = model_data, 
                ctmax ~ species * start_temp + (1|tube))

performance::check_model(temp.model)

summary(temp.model)

car::Anova(temp.model, type = "III")

```

```{r}

emmeans::emmeans(temp.model, ~species) %>% plot() + labs(x = "Marginal Mean (°C)",
                                                         y = "Species") + theme_matt()

```


```{r}
ggplot(trait_data, aes(x = species, y = ctmax)) + 
  geom_boxplot() + 
  geom_point() + 
  labs(x = "Species", 
       y = "CTmax (°C)") + 
  theme_matt()
```

```{r}

trait_data %>%  
  filter(!is.na(length)) %>% 
  ggplot(aes(x = length, y = ctmax, colour = species)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme_matt() + 
  theme(legend.position = "right")

```

## References
